### 2015-07-10 Paper_v2.tex
Updates
Latest Updates

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large Modeling and Analysis of Hormone and Mitogenic Signal Integration in Prostate Cancer}}
\vspace{0.05in}
{\par \centering \large{ Katharine V. Rogers, Joseph A. Wayman, Ryan Tasseff, Caitlin Gee, Timothy Chu, Kyriakos Tsahalis, Spencer Davis, Rohit Yallamraju, Gongcheng Lu, Matthew P. DeLisa, 
and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering \large{School of Chemical and Biomolecular Engineering}}
{\par \centering \large{Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Modeling signal transduction in prostate cancer}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Molecular Systems Biology}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853} 
{\par \centering Email: jdv27@cornell.edu} 
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
Prostate cancer is the most common cancer in men and the second leading cause of cancer related death in the United States. 
Androgens, such as testosterone, are required for prostate cancer growth. 
Androgen ablation in combination with radiation or chemotherapy remains the primary non-surgical treatment for androgen dependent prostate cancer. 
However, androgen ablation typically fails to permanently arrest cancer progression, often resulting in castration resistant prostate cancer (CRPC). 
CRPC is closely related to metastasis and decreased survival. 
In this study, we developed and analyzed a population of mathematical models describing growth factor and hormone signal integration in androgen dependent, intermediate and resistant prostate cancer cells. 
The model describes the integration of two simultaneous extracellular signaling inputs, namely, androgen and growth factors into a G1/S cell cycle checkpoint decision. 
Model parameters were identified from 43 studies in androgen dependent and resistant LNCaP cell lines. 
The model was validated by comparing simulations with an additional 29 data sets from LNCaP cell lines that were not used during training. 
Additionally, data from four drug trials was also used to evaluate the model’s performance.
Sensitivity analysis, conducted over an ensemble of prostate signaling models, suggested that in an androgen free environment general translation and transcription was more sensitive in androgen dependent cells, while in androgen independent cells the PI3K and MAPK pathway species were more sensitive. 
In a constant DHT environment sensitive species were conserved between the cell lines. 
These results suggest targeting the PI3K and MAPK pathways in addition to anti-androgen therapies as a treatment for CRPC.

@@

\subsection*{Estimating an ensemble of prostate signaling models.} 
We modeled the integration of growth factor, cell cycle and hormone signaling pathways in AD and CR LNCaP cells (Fig. \ref{fg:NetworkSchematic}). 
The signaling architecture was hand curated from over 80 primary literature sources in combination with biological databases.
The model equations were formulated as a system of ordinary differential equations (ODEs), where biochemical reaction rates were modeled using mass action kinetics.
ODEs and mass action kinetics are common modeling tools \citep{Wayman2013}, however, ODEs have the disadvantage of requiring estimates for unknown model parameters.
Many techniques have been developed to estimate ODE model parameters, often from noisy and sparse experimental data \citep{Moles:2003dn}.
Typically these identification problems are underdetermined, hence no unique parameter values can be estimated \citep{Villaverde:2014kl}.
Thus, instead of estimating a single yet highly uncertain parameter set, we estimated an ensemble of possible parameter sets using the Pareto Optimal Ensemble Techniques (POETs) algorithm \citep{Song2010}.
POETs uses a combination of simulated annealing and local optimization techniques coupled with Pareto optimality-based ranking to simultaneously optimize multiple objective functions.
Starting from an initial best fit set, we estimated the 1687 unknown model parameters (1674 kinetic parameters and 13 non-zero initial conditions) using 43 \textit{in vitro} data sets taken from AD, intermediate and CR LNCaP cells (Table~\ref{objective_table}). 
Each of the 43 training data sets was a separate objective in the multiobjective calculation. 
The training data were steady-state or dynamic immunoblots from which we extracted relative species abundance from their optical density profiles. 
POETs sampled well over a million possible parameter sets, from which we selected N = 5000 sets for further analysis. 
Over the ensemble, the coefficient of variation (CV) of the kinetic parameters spanned 0.5 - 5.8, with 33\% of the parameters having a CV of less than one (Fig. ZZ). 
As a control, we also performed simulations for R = 100 random parameter sets to compare against the ensemble generated by POETs.

@@

We also considered the sensitivity of CR cells following the application of the AR inhibitor enzalutamide in the presence of DHT (Fig. \ref{fg:Sensitivity}D). 
In an androgen environment with enzalutamide, species which were more fragile included cytosolic AR, cPAcP, and p21Cip1. 
As we would expect, AR species found in the nucleus and/or bound to coactivators, were more robust in the absence of enzalutamide. 
The top two percent of sensitive species with and without enzalutamide were conserved. 
In a CR cell, enzalutamide had no effect on the sensitivity of PI3K/AKT species as well as many MAPK species (ERK, Raf, and MEK). 
Next, we looked at the effect of enzalutamide on a CR cell in both a non-androgen and DHT environment (Fig. \ref{fg:Supp_Sensitivity}). 
More sensitive species in a non-androgen environment included dimerized HER2, ERK, and PAcP. 
Species which were more robust in the non-androgen environment included, AR activated by DHT, AKT, p70, and AR bound to HSP.
The results of our sensitivity analysis indicate that instead of inhibiting solely the AR pathway (enzalutamide), a combination therapy targeting the PI3K or MAPK pathways in addition to AR may be more effective.

@@

The results from our sensitivity analysis indicate that instead of inhibiting solely the AR pathway (enzalutamide), a combination therapy is necessary. 
To test this hypothesis we used the well characterized ADPC cell line LNCaP as well a LNCaP derived CRPC cell line C4-2 \cite{Thalmann1994}. 
Three inhibitors were used: the AR inhibitor MDV3100 (enzalutamide), the Raf kinase inhibitor sorafenib, and the PI3K inhibitor LY294002. 
In both cell lines, inhibition of either the AR or MAPK pathway appears to promote activation of the PI3K pathway, as seen by the increase in pAKT (S473) (Fig. \ref{fg:Experiment}A). 
The addition of the PI3K inhibitor, LY294002, alone or in combination diminishes PI3K activity. 
The inhibition of PI3K alone, led to an increase in AR expression in both LNCaP and C4-2 cell lines. 
Since, AR transcriptionally upregulates its own expression \cite{Lin2003}\cite{Grad1999}, this may indicate an increase in AR activity. 
The ribosomal protein pS6 was completely inhibited only in the presence of the PI3K inhibitor LY294002. 
Cell viability results show a large decrease in cell viability at 72 hrs in the dual inhibition cases as well as the triple inhibition case for both LNCaP and C4-2 cell lines (Fig. \ref{fg:Experiment}B). I
n both cell lines, MDV3100 (10 $\mu$M), has only a modest effect on cell viability versus control (DMSO). 
(Fig. \ref{fg:Experiment}C) shows cell viability of both cell lines in varying concentrations of inhibitors at 24 hrs.

\subsection*{Robustness analysis identified key regulators of prostate cancer.} 
Robustness analysis was conducted for 80 proteins to quantify the effects of amplifying or knocking down key model components in both AD and CR cells. 
Gene expression parameters were altered by a factor 10, 0.5, and 0 for knock-in, knock-down, or knock-out perturbations, respectively.
We calculated the effect of these perturbations on different protein markers, such as PSA, AR, and cyclin D. 
The results of the model suggest that an inhibition of either the PI3K pathway or the MAPK pathway in combination with an AR inhibitor as a possible therapy for CRPC. 
Sensitivity analysis revealed no change in the top sensitive species in the presence or absence of the AR inhibitor, enzalutamide. PI3K/AKT and MAPK species continued to fall in the top two percent of sensitive species. 
A study by Carver \textit{et al.} looked at dual inhibition of AR and PI3K signaling in LNCaP cells and in a Pten-deficient murine prostate cancer model \cite{Parker2013}. 
Using both the PI3K inhibitor, BEZ235, and the AR inhibitor, MDV3100 (enzalutamide), the group saw a drastic decrease in the total number of cells. 
Each inhibitor on it’s own had a much smaller effect on total cell number. 
They saw an increase in the cell death marker, c-PARP, in the dual inhibition case. 
The group hypothesized that AKT inhibition leads to increased AR signaling activity through increased protein concentrations of HER3. 
On the other hand, AR inhibition leads to increased AKT activity due to the down regulation of PHLPP, a protein phosphatase that regulates AKT. 
For simplicity, the HER3 pathway and also cell death were not included in the model. 
Dual knock-out studies of PI3K and AR in our model show no additive effect on any cell cycle proteins through the dual knock-out. 
The pathways in our model appear to be uncoupled and therefore no synergy is shown in the dual knock-out case. 
This could indicate that the combined decrease in cell population is entirely due to cell death. 
The Carver \textit{et al.} study did not consider cell cycle proteins or cell growth. 
Our model does show a decrease in cell cycle proteins in the PI3K knock-out as well as in the dual knock-out case. 
This result seems to be consistent with the decreased cell growth in the PI3K knock-out case which is not dependent on cell death, as c-PARP levels are low. 
The decrease in cell cycle proteins is due to a decrease in general translation, including free eIF4E levels and activated 40S ribosome subunit. 
The decrease in p70 (S6) activation due to inhibition of PI3K is shown in both the model and in the Carver \textit{et al.} study (Supplementary Figure), indicating this result is due to the PI3K pathway. 
It would be interesting to repeat the experiment looking at cell cycle proteins and also performing the experiment in CR LNCaP cells, instead of AD cells.

A knock-out of Raf, MEK or ERK showed an overall increase in cyclin D levels in CR cells (Fig. \ref{fg:Supp_Robustness}). 
This was unexpected and we saw a similar increase in cyclin D due to the knock-in of Raf, MEK or ERK. 
We found that individual ensemble members showed different response to a Raf knock-out, in both cyclin D concentration and PSA concentration. 
Of the 500 ensemble members, 126 members saw an increase in PSA concentration and 62 members saw an increase in cyclin D concentration due to the knock-out of Raf (Fig. \ref{fg:Robustness}). 
We saw three distinct regions: (1) increase in PSA concentration, (2) increase in cyclin D concentration, and (3) a decrease in both PSA and cyclin D. 
We explored the flux vectors of the outlying parameter sets to understand the mechanistic effect of Raf knock-out on PSA and cyclin D. 
Outlying parameter sets in region 1 displayed high activation of PI3K through HER2 signaling as well as high association of AP1 with AR. 
AP1 binds and suppresses AR transcriptional activity in LNCaP cells \cite{Sato1997}. 
Knocking out Raf lowered AP1 levels and, therefore, freed AR for increased transcription of PSA.
In region 2, parameter sets also had high activation of PI3K through HER2. 
They also had higher association of E2F with Rb and cyclin D1a with AR. 
Cyclin D levels in this region increase due to an increase in E2F levels caused by the Raf knock-out. 
Parameters in region 3 have high association of TOR. 
Interestingly, the drug sorafenib, a multi-kinase inhibitor that has activity against Raf, showed no measurable PSA decline in prostate cancer patients in clinical trials \cite{Dahut2008}. 
The robustness analysis showed that network perturbation can result in unexpected responses due to cell-to-cell heterogeneity in gene expression. 
These outlying cell types could be critical for understanding when designing drug targets and combination therapies.

@@

The results of the model suggest that an inhibition of either the PI3K pathway or the MAPK pathway in combination with an AR inhibitor as a possible therapy for CRPC. 
Sensitivity analysis revealed no change in the top sensitive species in the presence or absence of the AR inhibitor, enzalutamide. 
PI3K/AKT and MAPK species continued to fall in the top two percent of sensitive species. 
A study by Carver \textit{et al.} looked at dual inhibition of AR and PI3K signaling in LNCaP cells and in a Pten-deficient murine prostate cancer model \cite{Parker2013}. 
Using both the PI3K inhibitor, BEZ235, and the AR inhibitor, MDV3100 (enzalutamide), the group saw a drastic decrease in the total number of cells. 
Each inhibitor on it’s own had a much smaller effect on total cell number. They saw an increase in the cell death marker, c-PARP, in the dual inhibition case. 
The group hypothesized that AKT inhibition leads to increased AR signaling activity through increased protein concentrations of HER3. 
On the other hand AR inhibition leads to increased AKT activity due to the down regulation of PHLPP, a protein phosphatase that regulates AKT. 
For the simplicity of this model, the HER3 pathway and also cell death were not included in the model. 
Dual knock-out studies of PI3K and AR in our model show no additive effect on any cell cycle proteins through the dual knock-out. 
The pathways in our model appear to be uncoupled and therefore no synergy is shown in the dual knock-out case. 
This could indicate that the combined decrease in cell population is entirely due to cell death. 
The Carver \textit{et al.} study did not consider cell cycle proteins or cell growth. 
Our model does show a decrease in cell cycle proteins in the PI3K knock-out as well as in the dual knock-out case. 
This result seems to be consistent with the decreased cell growth in the PI3K knock-out case which is not dependent on cell death, as c-PARP levels are low. 
The decrease in cell cycle proteins is due to a decrease in general translation, including free eIF4E levels and activated 40S ribosome subunit. 
The decrease in p70 (S6) activation due to inhibition of PI3K is shown in both the model and in the Carver \textit{et al.} study (Supplementary Figure), indicating this result is due to the PI3K pathway. 
Our results confirm the Carver \textit{et al.} study in that a dual inhibition of AR and PI3K signaling led to a more prominent decrease in cell viability then each of the inhibitors alone. 
We extended the study to look at the addition of a third inhibitor, sorafenib, that inhibits Raf kinase in the MAPK pathway.

@@

\section*{Acknowledgements}
This study was supported by award number U54CA143876 from the National Cancer Institute.
The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Cancer Institute or the National Institutes of Health.
The authors acknowledge the financial support of the National Science Foundation, award no. DGE-1045513,for a GK12 training grant entitled "Grass Roots: Advancing Education in Renewable Energy and Cleaner Fuels through collaborative graduate fellow/teacher/grade-school student interactions."

@@

\begin{figure}\centering
\includegraphics[width=1.0\textwidth]{./figs/Fig_6_PI3K_AR_KO_3_Cases_old_new.pdf}
\caption{Robustness analysis of a population of CR prostate models with PI3K (blue), AR (green), and PI3K and AR (orange) knock-outs (N = 500). A log fold change of greater then zero implies that the concentration of the protein increased with the knock-out, while a log fold change of less then zero indicates that the concentration of protein decreased. A log of fold change equal to 0, shows no response due to the knock-out.
 A.,B.,C. Log robustness of cyclin E, PSA, and p70p versus Cyclin D for the three knock-out cases. A CR LNCaP cell was assumed for all knock-out cases.}
\label{fg:Supp_Dual_Knockout}
\end{figure}

@@

\begin{figure}\centering
\includegraphics[width=1.0\textwidth]{./figs/Supp_Figure_Robustness.pdf}
\caption{Robustness analysis of protein markers. Expression level of key proteins was altered by a factor of 2, 0.1, or 0 (knock-in, knock-down, or knock-out) and robustness coefficients were calculated for five key protein markers. Simulations shown were from CR cells, with indicated perturbation. Mean of 500 ensemble members is shown. }
\label{fg:Supp_Robustness}
\end{figure}

>>>>

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large Modeling and Analysis of Hormone and Mitogenic Signal Integration in Prostate Cancer}}
\vspace{0.05in}
{\par \centering \large{ Katharine V. Rogers, Joseph A. Wayman, Ryan Tasseff, Caitlin Gee, Matthew P. DeLisa, 
and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering \large{School of Chemical and Biomolecular Engineering}}
{\par \centering \large{Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Modeling signal transduction in prostate cancer}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Molecular Systems Biology}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853} 
{\par \centering Email: jdv27@cornell.edu} 
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
Prostate cancer is the most common cancer in men and the second leading cause of cancer related death in the United States. 
Androgens, such as testosterone, are required for prostate cancer growth. 
Androgen ablation in combination with radiation or chemotherapy remains the primary non-surgical treatment for androgen dependent prostate cancer. 
However, androgen ablation typically fails to permanently arrest cancer progression, often resulting in castration resistant prostate cancer (CRPC). 
CRPC is closely related to metastasis and decreased survival. 
In this study, we developed and analyzed a population of mathematical models describing growth factor and hormone signal integration in androgen dependent, intermediate and resistant prostate cancer cells. 
The model describes the integration of two simultaneous extracellular signaling inputs, namely, androgen and growth factors into a G1/S cell cycle checkpoint decision. 
Model parameters were identified from 43 studies in androgen dependent and resistant LNCaP cell lines. 
The model was validated by comparing simulations with an additional 29 data sets from LNCaP cell lines that were not used during training. 
Additionally, data from four drug trials was also used to evaluate the model’s performance.
Sensitivity analysis, conducted over an ensemble of prostate signaling models, suggested that in an androgen free environment general translation and transcription was more sensitive in androgen dependent cells, while in androgen independent cells the PI3K and MAPK pathway species were more sensitive. 
In a constant DHT environment sensitive species were conserved between the cell lines. 
These results suggest targeting the PI3K and MAPK pathways in addition to anti-androgen therapies as a treatment for CRPC.

@@

\subsection*{Estimating an ensemble of prostate signaling models.} 
We modeled the integration of growth factor, cell cycle and hormone signaling pathways in AD and CR LNCaP cells (Fig. \ref{fg:NetworkSchematic}). 
The signaling architecture was hand curated from over 80 primary literature sources in combination with biological databases.
The model equations were formulated as a system of ordinary differential equations (ODEs), where biochemical reaction rates were modeled using mass action kinetics.
ODEs and mass action kinetics are common modeling tools \citep{Wayman2013}, however, ODEs have the disadvantage of requiring estimates for unknown model parameters.
Many techniques have been developed to estimate ODE model parameters, often from noisy and sparse experimental data \citep{Moles:2003dn}.
Typically these identification problems are underdetermined, hence no unique parameter values can be estimated \citep{Villaverde:2014kl}.
Thus, instead of estimating a single yet highly uncertain parameter set, we estimated an ensemble of possible parameter sets using the Pareto Optimal Ensemble Techniques (POETs) algorithm \citep{Song2010}.
POETs uses a combination of simulated annealing and local optimization techniques coupled with Pareto optimality-based ranking to simultaneously optimize multiple objective functions.
Starting from an initial best fit set, we estimated the 1687 unknown model parameters (1674 kinetic parameters and 13 non-zero initial conditions) using 43 \textit{in vitro} data sets taken from AD, intermediate and CR LNCaP cells (Table~\ref{objective_table}). 
Each of the 43 training data sets was a separate objective in the multiobjective calculation. 
The training data were steady-state or dynamic immunoblots from which we extracted relative species abundance from their optical density profiles. 
POETs sampled well over a million possible parameter sets, from which we selected N = 5000 sets for further analysis. 
Over the ensemble, the coefficient of variation (CV) of the kinetic parameters spanned 0.59 - 5.8, with 33\% of the parameters having a CV of less than one (Fig. \ref{fg:Supp_CV}). 
As a control, we also performed simulations for R = 100 random parameter sets to compare against the ensemble generated by POETs.

@@

We also considered the sensitivity of CR cells following the application of the AR inhibitor enzalutamide in the presence of DHT (Fig. \ref{fg:Sensitivity}D). 
Species which were more sensitive in an androgen environment with enzalutamide included cytosolic AR, cPAcP, and p21Cip1. 
As we would expect, AR species found in the nucleus and/or bound to coactivators, were more robust in the absence of enzalutamide. 
The top two percent of sensitive species with and without enzalutamide were conserved. 
In a CR cell, enzalutamide had no effect on the sensitivity of PI3K/AKT species as well as many MAPK species (ERK, Raf, and MEK). 
Next, we looked at the effect of enzalutamide on a CR cell in both a non-androgen and DHT environment (Fig. \ref{fg:Supp_Sensitivity}). 
More sensitive species in a non-androgen environment included dimerized HER2, ERK, and PAcP. 
Species which were more robust in the non-androgen environment included, AR activated by DHT, AKT, p70, and AR bound to HSP.
The results of our sensitivity analysis indicate that instead of inhibiting solely the AR pathway (enzalutamide), a combination therapy targeting the PI3K or MAPK pathways in addition to AR may be more effective.

@@

The results from our sensitivity analysis indicate that instead of inhibiting solely the AR pathway (enzalutamide), a combination therapy is necessary. 
To test this hypothesis we used the well characterized ADPC cell line LNCaP as well a LNCaP derived CRPC cell line C4-2 \cite{Thalmann1994}. 
Three inhibitors were used: the AR inhibitor MDV3100 (enzalutamide), the Raf kinase inhibitor sorafenib, and the PI3K inhibitor LY294002. 
In both cell lines, inhibition of either the AR or MAPK pathway appears to promote activation of the PI3K pathway, as seen by the increase in pAKT (S473) (Fig. \ref{fg:Experiment}A). 
The addition of the PI3K inhibitor, LY294002, alone or in combination diminishes PI3K activity. 
The inhibition of PI3K alone, led to an increase in AR expression in both LNCaP and C4-2 cell lines. 
Since, AR transcriptionally upregulates its own expression \cite{Lin2003}\cite{Grad1999}, this may indicate an increase in AR activity. 
The ribosomal protein pS6 was completely inhibited only in the presence of the PI3K inhibitor LY294002. 
Cell viability results show a large decrease in cell viability at 72 hrs in the dual inhibition cases as well as the triple inhibition case for both LNCaP and C4-2 cell lines (Fig. \ref{fg:Experiment}B). 
In both cell lines, MDV3100 (10 $\mu$M), has only a modest effect on cell viability versus control (DMSO). 
Figure \ref{fg:Experiment}C shows cell viability of both cell lines in varying concentrations of inhibitors at 24 hrs.

\subsection*{Robustness analysis identified key regulators of prostate cancer.} 
Robustness analysis was conducted for 80 proteins to quantify the effects of amplifying or knocking down key model components in both AD and CR cells. 
Gene expression parameters were altered by a factor 10, 0.5, and 0 for knock-in, knock-down, or knock-out perturbations, respectively.
We calculated the effect of these perturbations on different protein markers, such as PSA, AR, and cyclin D. 
A knock-out of Raf, MEK or ERK showed an overall increase in cyclin D levels in CR cells (Fig. \ref{fg:Supp_Robustness}). 
This was unexpected and we saw a similar increase in cyclin D due to the knock-in of Raf, MEK or ERK. 
We found that individual ensemble members showed different response to a Raf knock-out, in both cyclin D concentration and PSA concentration. 
Of the 500 ensemble members, 126 members saw an increase in PSA concentration and 62 members saw an increase in cyclin D concentration due to the knock-out of Raf (Fig. \ref{fg:Robustness}). 
We saw three distinct regions: (1) increase in PSA concentration, (2) increase in cyclin D concentration, and (3) a decrease in both PSA and cyclin D. 
We explored the flux vectors of the outlying parameter sets to understand the mechanistic effect of Raf knock-out on PSA and cyclin D. 
Outlying parameter sets in region 1 displayed high activation of PI3K through HER2 signaling as well as high association of AP1 with AR. 
AP1 binds and suppresses AR transcriptional activity in LNCaP cells \cite{Sato1997}. 
Knocking out Raf lowered AP1 levels and, therefore, freed AR for increased transcription of PSA.
In region 2, parameter sets also had high activation of PI3K through HER2. 
They also had higher association of E2F with Rb and cyclin D1a with AR. 
Cyclin D levels in this region increase due to an increase in E2F levels caused by the Raf knock-out. 
Parameters in region 3 have high association of TOR. 
Interestingly, the drug sorafenib, a multi-kinase inhibitor that has activity against Raf, showed no measurable PSA decline in prostate cancer patients in clinical trials \cite{Dahut2008}. 
The robustness analysis showed that network perturbation can result in unexpected responses due to cell-to-cell heterogeneity in gene expression. 
These outlying cell types could be critical for understanding when designing drug targets and combination therapies.

%The results of the model suggest that an inhibition of either the PI3K pathway or the MAPK pathway in combination with an AR inhibitor as a possible therapy for CRPC. 
%Sensitivity analysis revealed no change in the top sensitive species in the presence or absence of the AR inhibitor, enzalutamide. PI3K/AKT and MAPK species continued to fall in the top two percent of sensitive species. 
%A study by Carver \textit{et al.} looked at dual inhibition of AR and PI3K signaling in LNCaP cells and in a Pten-deficient murine prostate cancer model \cite{Parker2013}. 
%Using both the PI3K inhibitor, BEZ235, and the AR inhibitor, MDV3100 (enzalutamide), the group saw a drastic decrease in the total number of cells. 
%Each inhibitor on it’s own had a much smaller effect on total cell number. 
%They saw an increase in the cell death marker, c-PARP, in the dual inhibition case. 
%The group hypothesized that AKT inhibition leads to increased AR signaling activity through increased protein concentrations of HER3. 
%On the other hand, AR inhibition leads to increased AKT activity due to the down regulation of PHLPP, a protein phosphatase that regulates AKT. 
%For simplicity, the HER3 pathway and also cell death were not included in the model. 
%Dual knock-out studies of PI3K and AR in our model show no additive effect on any cell cycle proteins through the dual knock-out. 
%The pathways in our model appear to be uncoupled and therefore no synergy is shown in the dual knock-out case. 
%This could indicate that the combined decrease in cell population is entirely due to cell death. 
%The Carver \textit{et al.} study did not consider cell cycle proteins or cell growth. 
%Our model does show a decrease in cell cycle proteins in the PI3K knock-out as well as in the dual knock-out case. 
%This result seems to be consistent with the decreased cell growth in the PI3K knock-out case which is not dependent on cell death, as c-PARP levels are low. 
%The decrease in cell cycle proteins is due to a decrease in general translation, including free eIF4E levels and activated 40S ribosome subunit. 
%The decrease in p70 (S6) activation due to inhibition of PI3K is shown in both the model and in the Carver \textit{et al.} study (Supplementary Figure), indicating this result is due to the PI3K pathway. 
%It would be interesting to repeat the experiment looking at cell cycle proteins and also performing the experiment in CR LNCaP cells, instead of AD cells.

@@

The results of the model suggest that an inhibition of either the PI3K pathway or the MAPK pathway in combination with an AR inhibitor as a possible therapy for CRPC. 
Sensitivity analysis revealed no change in the top sensitive species in the presence or absence of the AR inhibitor, enzalutamide. 
PI3K/AKT and MAPK species continued to fall in the top two percent of sensitive species. 
A study by Carver \textit{et al.} looked at dual inhibition of AR and PI3K signaling in LNCaP cells and in a Pten-deficient murine prostate cancer model \cite{Parker2013}. 
Using both the PI3K inhibitor, BEZ235, and the AR inhibitor, MDV3100 (enzalutamide), the group saw a drastic decrease in the total number of cells. 
Each inhibitor on it’s own had a much smaller effect on total cell number. They saw an increase in the cell death marker, c-PARP, in the dual inhibition case. 
The group hypothesized that AKT inhibition leads to increased AR signaling activity through increased protein concentrations of HER3. 
On the other hand AR inhibition leads to increased AKT activity due to the down regulation of PHLPP, a protein phosphatase that regulates AKT. 
For the simplicity of this model, the HER3 pathway and also cell death were not included in the model. 
Dual knock-out studies of PI3K and AR in our model show no additive effect on any cell cycle proteins through the dual knock-out. 
The pathways in our model appear to be uncoupled and therefore no synergy is shown in the dual knock-out case. 
This could indicate that the combined decrease in cell population is entirely due to cell death. 
The Carver \textit{et al.} study did not consider cell cycle proteins or cell growth. 
Our model does show a decrease in cell cycle proteins in the PI3K knock-out as well as in the dual knock-out case. 
This result seems to be consistent with the decreased cell growth in the PI3K knock-out case which is not dependent on cell death, as c-PARP levels are low. 
The decrease in cell cycle proteins is due to a decrease in general translation, including free eIF4E levels and activated 40S ribosome subunit. 
The decrease in p70 (S6) activation due to inhibition of PI3K is shown in both the model and in the Carver \textit{et al.} study ((Fig. \ref{fg:Supp_Figure_KO_new.pdf})), indicating this result is due to the PI3K pathway. 
Our results confirm the Carver \textit{et al.} study in that a dual inhibition of AR and PI3K signaling led to a more prominent decrease in cell viability then each of the inhibitors alone. 
We extended the study to look at the addition of a third inhibitor, sorafenib, that inhibits Raf kinase in the MAPK pathway.

@@

\section*{Acknowledgements}
This study was supported by award number U54CA143876 from the National Cancer Institute.
The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Cancer Institute or the National Institutes of Health.
The authors acknowledge the financial support of the National Science Foundation, award no. DGE-1045513,for a GK12 training grant entitled "Grass Roots: Advancing Education in Renewable Energy and Cleaner Fuels through collaborative graduate fellow/teacher/grade-school student interactions."
We would like to thank Timothy Chu, Kyriakos Tsahalis, Spencer Davis, Rohit Yallamraju, Gongcheng Lu, and Yalin Zhao.

@@

\begin{figure}\centering
\includegraphics[width=1.0\textwidth]{./figs/Supplementary_CV_Subset_Total}
\caption{Coefficient of variation (CV) of model parameters estimated using POETs. The solid line denotes the mean CV calculated over the entire ensemble (N = 5000). The points denote the mean CV of the 500 ensemble members used for sensitivity and robustness calculations. Over the ensemble, the coefficient of variation (CV) of the kinetic parameters spanned 0.59 - 5.8, with 33\% of the parameters having a CV of less than one.}
\label{fg:Supp_CV}
\end{figure}

\begin{figure}\centering
\includegraphics[width=1.0\textwidth]{./figs/Supp_Figure_NewModelvsOldModel}
\caption{Blind model predictions for the ensemble with the original and updated model (EGFR and HER2 heterodimer). A,B. Time course data for HER phosphorylation due to a stimulus of 1.6 nM EGF (LNCaP C33, P7) for the old and new model, respectively. Dark grey shows only parameters improved by the updated model (N=2870) while light grey show all parameter sets (N=5000). C,D. Shc phosphorylation levels at 16 hrs in the presence and absence of 1.6 nM EGF (LNCaP C33, P29) for the old and new model, respectively. Light grey denotes experimental data, mid grey denotes simulation results for all parameters (N=5000), and black denotes only parameters improved by the updated model (N=3142).}
\label{fg:Exps_NewModelvsOldModel}
\end{figure}

\begin{figure}\centering
\includegraphics[width=1.0\textwidth]{./figs/Fig_6_PI3K_AR_KO_3_Cases.pdf}
\caption{Robustness analysis of a population of CR prostate models with PI3K (blue), AR (green), and PI3K and AR (orange) knock-outs (N = 500). A log fold change of greater then zero implies that the concentration of the protein increased with the knock-out, while a log fold change of less then zero indicates that the concentration of protein decreased. A log of fold change equal to 0, shows no response due to the knock-out.
 A.,B.,C. Log robustness of cyclin E, PSA, and p70p versus Cyclin D for the three knock-out cases. A CR LNCaP cell was assumed for all knock-out cases.}
\label{fg:Supp_Dual_Knockout}
\end{figure}

@@

\begin{figure}\centering
\includegraphics[width=1.0\textwidth]{./figs/Supp_Figure_Robustness.pdf}
\caption{Robustness analysis of protein markers. Expression level of key proteins was altered by a factor of 2, 0.1, or 0 (knock-in, knock-down, or knock-out) and robustness coefficients were calculated for five key protein markers. Simulations shown were from CR cells, with indicated perturbation. Mean of 500 ensemble members is shown.}
\label{fg:Supp_Robustness}
\end{figure}
